NLTK: The Natural Language Toolkit

### 2002-05-06 technical/acl2002/acl02.tex
Several minor editorial corrections
svn/trunk@551

%
% Submission for ``Effective Tools And Methodologies For Teaching
% Natural Language Processing And Computational Linguistics'' workshop
% at ACL 2002
%
% Authors: Edward Loper and Steven Bird
% 
% $Id$
%
\documentclass[11pt]{article}
\usepackage{acl2002,url,alltt,epsfig}
\title{NLTK: The Natural Language Toolkit}
\author{
Edward Loper and Steven Bird\\
Department of Computer and Information Science and Linguistic Data Consortium\\
University of Pennsylvania, Philadelphia, PA 19104-6389, USA
}
\date{}

@@

\begin{abstract}
NLTK, the Natural Language Toolkit, is a suite of program modules,
tutorials and problem sets, providing ready-to-use computational
linguistics courseware.  NLTK covers symbolic and statistical natural
language processing, and is interfaced to annotated corpora.  Students
augment and replace existing components, learning structured
programming by example, and manipulating sophisticated models from the
outset.
\end{abstract}

@@

% [EL] ``functionality'' isn't a great word, here.
Teachers of introductory courses on computational linguistics are
often faced with the challenge of setting up a practical programming
component for student assignments and projects.  This is a difficult
task because different computational linguistics domains require a
variety of different data structures and functionality; and because a
diverse range of topics may need to be included in the syllabus.

A widespread practice is to employ multiple programming languages,
where each language provides native data structures and functions that
are a good fit for the task at hand.  For example, a course might use
Prolog for parsing, Perl for corpus processing, and a finite-state
toolkit such for morphological analysis.  By relying on the built-in
features of various languages, the teacher avoids having to develop a
lot of software infrastructure.

An unfortunate consequence is that a significant part of the course
must be devoted to teaching these languages.  Further, many
interesting projects span a variety of domains, and would require that
multiple languages be bridged.  For example, a student project that
involved syntactic parsing of corpus data from a morphologically rich
language might involve all three of the languages mentioned above:
Perl for string processing; a finite state toolkit for morphological
analysis; and Prolog for parsing.
% [EL] This was more precise, but harder to read/parse:
%    (Perl for file I/O, format conversions and
%    overall program control, with calls out to a finite state toolkit for
%    morphological analysis and to a Prolog engine for parsing).  
It is clear that these considerable overheads and shortcomings warrant
a fresh approach.

% [EL] ``difficulties'' doesn't seem like the right word here?
% Maybe ``challenging requirements''?
Apart from the practical component, computational linguistics courses
may also depend on software for in-class demonstrations.  This context
calls for highly interactive graphical user interfaces, making it
possible to view program state (e.g. the chart of a chart parser),
observe program execution step-by-step (e.g. execution of a
finite-state machine), and even make minor modifications to programs
in response to ``what if'' questions from the class.  Because of these
difficulties it is common to avoid live demonstrations, and keep
classes for theoretical presentations only.  Apart from being dull,
this approach leaves students to solve important practical problems on
their own, or to deal with them less efficiently in office hours.

@@

% Overview
% [EL] Given that we're short on space, I decided that this wasn't
% really necessary for a 6-page paper.  If we do include it, it needs
% to be completed (I only went up through sec:uses).
%%Section \ref{sec:python} discusses why we decided to implement the
%%toolkit using Python.  Section \ref{sec:criteria} describes several
%%important considerations and desiderata that guided the design and
%%implementation of the toolkit.  Sections \ref{sec:modules} and
%%\ref{sec:documentation} describe the current contents of the toolkit.
%%Section \ref{sec:uses} explains how the toolkit can be used for both
%%assignments and class demonstrations.

% [EL] This was too ambiguous for me (e.g., we discuss Python)
%All materials discussed here are available under an open source
%license from \url{nltk.sf.net}.  NLTK runs on most platforms,
%including Windows, OS X, Linux, and UNIX.
The Natural Language Toolkit is available under an open source license
from \mbox{\url{http://nltk.sf.net/}}.  NLTK runs on most platforms, including
Windows, OS X, Linux, and UNIX.

@@

% [EL] ``punitive'' like a strange word.
% [EL] I couldn't find a REF for ``executable pseudocode'' -- closest
% I got was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
In surveying the available languages, we believe that Python offers an
especially good fit to the above requirements.  Python is an
object-oriented scripting language developed by Guido van Rossum
and available on all platforms \cite{python}.  Python offers a
shallow learning curve; it was designed to be easily learnt by
children \cite{rossum99}.  As an interpreted language, Python is
suitable for rapid prototyping.  Python code is exceptionally
readable, and it has been praised as ``executable pseudocode.''
Python is an object-oriented language, but not punitively so, and it
is easy to encapsulate data and methods inside Python classes.
Finally, Python has an interface to the ``Tk'' graphics toolkit
\cite{tkinter}, and writing graphical interfaces is straightforward.

@@

NLTK can be used to create student assignments of varying difficulty
and scope.  
% Use a module
In the simplest assignments, students experiment with an existing
module.  The wide variety of existing modules provide many opportunities
for creating these simple assignments.
% Edit/extend a module.
Once students become more familiar with the toolkit, they can be asked
to make minor changes or extensions to an existing module.
% Develop a new module.
A more challenging task is to develop a new module.  Here, NLTK
provides some useful starting points: predefined interfaces and data
structures, and existing modules that implement the same interface.
% Advanced projects..
Advanced projects typically involve the development of entirely new
functionality for a previously unsupported NLP task, or the
development of a complete system out of existing and new modules.

@@

NLTK provides graphical tools that can be used in class demonstrations
to help explain basic NLP concepts and algorithms.  These interactive
tools can be used to display relevant data structures and to show the
step-by-step execution of algorithms.  Both data structures and
control flow can be easily modified during the demonstration, in
response questions from the class.

@@

@@

% Do we want to mention that there were 40h of lecture time?
% Should we be more explicit about who took the class?
We used NLTK as a basis for the assignments and student projects in
CIS-530, an introductory computational linguistics class taught at the
University of Pennsylvania.  CIS-530 is a graduate level class,
although some advanced undergraduates were also enrolled.  Most
students had a background in either computer science or linguistics
(but not both).  Students were required to complete five assignments,
two exams, and a final project.  All class materials are available
from the course website \mbox{\url{http://www.cis.upenn.edu/~cis530/}}.

@@

We did encounter a few difficulties during the semester.  One problem
was finding large clean corpora that the students could use for their
assignments.  Several of the students needed assistance finding
suitable corpora for their final projects.  Another issue was the fact
that we were actively developing NLTK during the semester; some
modules were only completed one or two weeks before the students used
them.  As a result, students who worked at home needed to download new
versions of the toolkit several times throughout the semester.
Luckily, Python has extensive support for installation scripts, which
made these upgrades much simpler.  The students encountered a couple
of bugs in the toolkit, but none were serious, and all were quickly
corrected.

@@

%Overview of the paper:
% [EL] Pick better 3 adjectives :)
NLTK provides a simple, extensible, uniform framework for student
assignments and class demonstrations.  It is well documented, easy to
learn, and simple to use.  We hope that NLTK will allow computational
linguistics classes to include more hands-on experience with using and
building NLP components and systems.

>>>>

%
% Submission for ``Effective Tools And Methodologies For Teaching
% Natural Language Processing And Computational Linguistics'' workshop
% at ACL 2002
%
% Authors: Edward Loper and Steven Bird
% 
% $Id$
%
\documentclass[11pt]{article}
\usepackage{acl2002,url,alltt,epsfig}
\title{NLTK: The Natural Language Toolkit}
\author{
Edward Loper and Steven Bird\\
Department of Computer and Information Science \\
University of Pennsylvania, Philadelphia, PA 19104-6389, USA
}
\date{}

@@

\begin{abstract}
NLTK, the Natural Language Toolkit, is a suite of open source program modules,
tutorials and problem sets, providing ready-to-use computational
linguistics courseware.  NLTK covers symbolic and statistical natural
language processing, and is interfaced to annotated corpora.  Students
augment and replace existing components, learning structured
programming by example, and manipulating sophisticated models from the
outset.
\end{abstract}

@@

Teachers of introductory courses on computational linguistics are
often faced with the challenge of setting up a practical programming
component for student assignments and projects.  This is a difficult
task because different computational linguistics domains require a
variety of different data structures and functions; and because a
diverse range of topics may need to be included in the syllabus.

A widespread practice is to employ multiple programming languages,
where each language provides native data structures and functions that
are a good fit for the task at hand.  For example, a course might use
Prolog for parsing, Perl for corpus processing, and a finite-state
toolkit for morphological analysis.  By relying on the built-in
features of various languages, the teacher avoids having to develop a
lot of software infrastructure.

An unfortunate consequence is that a significant part of such courses
must be devoted to teaching programming languages.  Further, many
interesting projects span a variety of domains, and would require that
multiple languages be bridged.  For example, a student project that
involved syntactic parsing of corpus data from a morphologically rich
language might involve all three of the languages mentioned above:
Perl for string processing; a finite state toolkit for morphological
analysis; and Prolog for parsing.
It is clear that these considerable overheads and shortcomings warrant
a fresh approach.

Apart from the practical component, computational linguistics courses
may also depend on software for in-class demonstrations.  This context
calls for highly interactive graphical user interfaces, making it
possible to view program state (e.g. the chart of a chart parser),
observe program execution step-by-step (e.g. execution of a
finite-state machine), and even make minor modifications to programs
in response to ``what if'' questions from the class.  Because of these
difficulties it is common to avoid live demonstrations, and keep
classes for theoretical presentations only.  Apart from being dull,
this approach leaves students to solve important practical problems on
their own, or to deal with them less efficiently in office hours.

@@

The Natural Language Toolkit is available under an open source license
from \mbox{\url{http://nltk.sf.net/}}.  NLTK runs on most platforms, including
Windows, OS X, Linux, and Unix.

@@

% [EL] ``punitive'' like a strange word.
% [EL] I couldn't find a REF for ``executable pseudocode'' -- closest
% I got was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
In surveying the available languages, we believe that Python offers an
especially good fit to the above requirements.  Python is an
object-oriented scripting language developed by Guido van Rossum
and available on all platforms \cite{python}.  Python offers a
shallow learning curve; it was designed to be easily learnt by
children \cite{rossum99}.  As an interpreted language, Python is
suitable for rapid prototyping.  Python code is exceptionally
readable, and it has been praised as ``executable pseudocode.''
Python is an object-oriented language, but not punitively so, and it
is easy to encapsulate data and methods inside Python classes.
Finally, Python has an interface to the Tk graphics toolkit
\cite{tkinter}, and writing graphical interfaces is straightforward.

@@

NLTK can be used to create student assignments of varying difficulty
and scope.  
% Use a module
In the simplest assignments, students experiment with an existing
module.  The wide variety of existing modules provide many opportunities
for creating these simple assignments.
% Edit/extend a module.
Once students become more familiar with the toolkit, they can be asked
to make minor changes or extensions to an existing module.
% Develop a new module.
A more challenging task is to develop a new module.  Here, NLTK
provides some useful starting points: predefined interfaces and data
structures, and existing modules that implement the same interface.

@@

NLTK provides graphical tools that can be used in class demonstrations
to help explain basic NLP concepts and algorithms.  These interactive
tools can be used to display relevant data structures and to show the
step-by-step execution of algorithms.  Both data structures and
control flow can be easily modified during the demonstration, in
response to questions from the class.

@@

\subsection{Advanced projects}

Advanced projects typically involve the development of entirely new
functionality for a previously unsupported NLP task, or the
development of a complete system out of existing and new modules.

@@

% Do we want to mention that there were 40h of lecture time?
% Should we be more explicit about who took the class?
We used NLTK as a basis for the assignments and student projects in
CIS-530, an introductory computational linguistics class taught at the
University of Pennsylvania.  CIS-530 is a graduate level class,
although some advanced undergraduates were also enrolled.  Most
students had a background in either computer science or linguistics
(and occasionally both).  Students were required to complete five assignments,
two exams, and a final project.  All class materials are available
from the course website \mbox{\url{http://www.cis.upenn.edu/~cis530/}}.

@@

We did encounter a few difficulties during the semester.  One problem
was finding large clean corpora that the students could use for their
assignments.  Several of the students needed assistance finding
suitable corpora for their final projects.  Another issue was the fact
that we were actively developing NLTK during the semester; some
modules were only completed one or two weeks before the students used
them.  As a result, students who worked at home needed to download new
versions of the toolkit several times throughout the semester.
Luckily, Python has extensive support for installation scripts, which
made these upgrades simple.  The students encountered a couple
of bugs in the toolkit, but none were serious, and all were quickly
corrected.

@@

%Overview of the paper:
% [EL] Pick better 3 adjectives :)
NLTK provides a simple, extensible, uniform framework for
assignments, projects, and class demonstrations.  It is well documented, easy to
learn, and simple to use.  We hope that NLTK will allow computational
linguistics classes to include more hands-on experience with using and
building NLP components and systems.
